The Fates of Merging Supermassive Black Holes and A Proposal for a New Class of X-Ray Sources

### 2019-11-08 paper.tex
Edits from Jerry.  Also, last push before switching to MNRAS template.

\title{Supermassive Black Hole Dynamics: A Proposal for a New Class of X-Ray Sources}

@@

\begin{abstract}
We perform N-body simulations on 11 galaxies extracted from a cosmological simulation of hierarchical structure formation with total mass in the range $10^{12} \msun < M < 10^{13} \msun$ from $4 \ge z \ge 0$.  After mergers, we track the dynamical evolution of the orbiting black holes (BHs) around their host central BHs.  Of the 86 orbiting BHs, we find that 36 merge with their host central BHs, 13 are ejected from their respective host galaxies, and 37 are still orbiting at the end of the simulations.  Across all galaxies, 33 BHs are kicked to a higher orbit after close interactions with the central BHs, after which only one of them merged with their hosts.  These orbiting BHs should be detectable by their anomalous (not LMXRB) spectra.  The x-ray luminosities of the orbiting massive BHs at z=0 are in the range $10^{25}-10^{43}$ $erg\ s^{-1}$, with median value $10^{32}$ $erg\ s^{-1}$.
\end{abstract}

@@

\section{Introduction}\label{sec:introduction}
What happens to the massive black holes centrally located in galaxies when those galaxies merge with others?  There are three possibilities: they merge, they are ejected, or they remain in orbit in the merged galaxy.  Combinations of these options are also possible with the spin induced kick at merger leading to an ejection or extended orbit of the merged black hole.  We now know from pulsar timing measurements (ADD REFERENCES) that the gravitational wave background is probably too low for most galaxy mergers (mass-weighted) (but see \cite{2018NatCo...9..573M}) to lead to BH mergers, but the outcome remains uncertain.  We will argue in this paper that a very common outcome--especially for the lower mass black holes--is for the injected objects to remain as orbiting X-ray sources in normal massive galaxies waiting to be identified by current observational techniques.

There is no "final parsec problem," and the opposite problem of too many BH mergers is also avoided when it is realized that we are dealing with a small N-body system rather than a series of two-body problems.

How common are mergers of massive black holes?  For the galaxies themselves we know that minor mergers are frequent from the observed (\citet{2010ApJ...718L..73V}, \citet{2008ApJ...677L...5V}, \citet{2019MNRAS.484..595M}) evolution of the size and mass of these systems, but assuming that all galaxy mergers lead to SMBH mergers overpredicts the observations of gravitational waves from the pulsar timing arrays (PTAs) (\citet{2008MNRAS.390..192S}, \citet{2009MNRAS.394.2255S}, \citet{2013MNRAS.433L...1S}, \citet{2014ApJ...789..156M}, \citet{2015ApJ...799..178K}, \citet{2018ApJ...856...42S}, \citet{2018arXiv180403143I}).  However, we do not often see multiple SMBHs at the centers of massive ellipticals.  So, what does happen?  One paper (\citet{2018MNRAS.473.3410R}) has indicated that dynamical interactions among multiple orbiting black holes, which will eject a non-negligible fraction of the mass, may solve this problem.  The present paper also addresses this purported dynamical solution, focusing attention on the large fraction of lower mass black holes that remain to be detected as they orbit in massive galaxies.  What we argue in this paper is that, as a result of the N-body interactions among the infalling black holes, some are ejected, some remain in extended orbits and some (the few most massive ones) do in fact merge, but these do not exceed the PTA limits, and the orbiting ones should be detectable via their ability to accrete gas and emit radiation.

@@

In cosmological simulations it is common to overproduce stellar mass, with a general over-efficiency in the range of 2-4 times those measured in observations (\citet{1996ApJS..105...19K}, \citet{2010MNRAS.404.1111G}, \citet{2010ApJ...725.2312O}).  One typical reason cited for this overproduction is a lack or underestimation of supernova feedback effects in the form of thermal and/or kinetic energy.  However, even when thermal energy from supernova feedback is considered, the effects may be reduced due to the energy being quickly radiated away by rapidly cooling surrounding gas (\citet{1996ApJS..105...19K}).  A second common reason for overproduction of stars in cosmological simulations is a lack of modeling for AGN feedback, which may be responsible for a phenomenon called the "cooling flow paradox" (\citet{2001MNRAS.321L..20F}) that suppresses star formation.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{plots/stellar_to_halo_ratio.png}
\caption{Approximation of Figure 11 from \citet{2018AstL...44....8K}.  Purple dots shown is example of our rescaled stellar masses}
\label{fig:stellar1}
\end{center}
\end{figure}

@@

\subsubsection{Kicked, Ejected and Orbiting Black Holes}\label{sec:kicked_orbiting}
Of the 37 orbiting black holes at $z=0$, 18 of them were kicked to higher orbits at some point in their traversal of their host galaxy.  If one counts the 13 ejected black holes, then of the 50 SMBHs that had not eventually merged with their host, 32 of them were kicked out of the galactic core as a result of three-body interactions.  If they didn't merge, it wasn't for a lack of effort.

@@

\subsection{Central Black Hole Influence}
Within the influence radius of the central black hole, which is defined as the radius at which the total mass of the galaxy within that radius (stars + dark matter) equals the mass of the central black hole, the gas density is much higher than otherwise (NEED REFERENCES?).  {WHAT DO WE SAY ABOUT THE VELOCITY FOLLOWING THE KEPLER PROFILE WITHIN THE INFLUENCE RADIUS?}.  Starting with the equation for hydrostatic equilibrium, 
\begin{equation}
    C_{s}^2\frac{dln\rho}{dr}= -\frac{GM(r)}{r^2},
\end{equation}
we can integrate with respect to r, leading to 
\begin{equation}
    ln\frac{\rho}{\rho_1} = \frac{GM}{C_{s}^2}\left(\frac{1}{r} - \frac{1}{r_1}\right),
\end{equation}
where $\rho$ is the gas density at the location of the wandering black hole, $\rho_1$ is the density at the radius where the central black hole mass equals the galaxy mass, $r$ is the distance of the wandering black hole from the center, and $r_1$ is the radius where the central black hole mass equals the galaxy mass.

@@

\section{Conclusions}\label{sec:conclusions}

>>>>

\title{The Fates of Merging Supermassive Black Holes and A Proposal for a New Class of X-Ray Sources}

@@

\begin{abstract}
We perform N-body simulations on 11 galaxies extracted from a cosmological simulation of hierarchical structure formation with total mass in the range $10^{12} \msun < M < 10^{13} \msun$ from $4 \ge z \ge 0$.  After mergers, we track the dynamical evolution of the orbiting black holes (BHs) around their host central BHs.  Of the 86 orbiting BHs, we find that 36 merge with their host central BHs, 13 are ejected from their respective host galaxies, and 37 are still orbiting at the end of the simulations.  Across all galaxies, 33 BHs are kicked to a higher orbit after close interactions with the central BHs, after which only one of them merged with their hosts.  These orbiting BHs should be detectable by their anomalous (not LMXRB) spectra.  The x-ray luminosities of the orbiting massive BHs at z=0 are in the range $10^{25}-10^{43}$ $erg\ s^{-1}$, with an undetectable median value $10^{32}$ $erg\ s^{-1}$.  However, the most luminous $\sim$$5\%$ should be detectable.
\end{abstract}

@@

\section{Introduction}\label{sec:introduction}
What happens to the massive black holes centrally located in galaxies when those galaxies merge with others?  There are three possibilities: they merge, they are ejected, or they remain in orbit in the merged galaxy.  Combinations of these options are also possible with the spin induced kick at merger leading to an ejection or extended orbit of the merged black hole.  We now know from pulsar timing measurements (\citet{2008MNRAS.390..192S}, \citet{2018ApJ...856...42S}, \citet{2016APS..APRR18003T}) that the gravitational wave background is probably too low for most galaxy mergers (mass-weighted) (but see \cite{2018NatCo...9..573M}) to lead to BH mergers, but the outcome remains uncertain.  We will argue in this paper that a very common outcome--especially for the lower mass black holes--is for the injected objects to remain as orbiting X-ray sources in normal massive galaxies waiting to be identified by current observational techniques.

There is no "final parsec problem," and the opposite problem of too many BH mergers is also avoided when it is realized that we are dealing with a small N-body system rather than a series of two-body problems (cf \citet{2018MNRAS.473.3410R}).

How common are mergers of massive black holes?  For the galaxies themselves we know that minor mergers are frequent from the observed (\citet{2010ApJ...718L..73V}, \citet{2008ApJ...677L...5V}, \citet{2019MNRAS.484..595M}) evolution of the size and mass of these systems, but assuming that all galaxy mergers lead to SMBH mergers overpredicts the observations of gravitational waves from the pulsar timing arrays (PTAs) (\citet{2008MNRAS.390..192S}, \citet{2009MNRAS.394.2255S}, \citet{2013MNRAS.433L...1S}, \citet{2014ApJ...789..156M}, \citet{2015ApJ...799..178K}, \citet{2018ApJ...856...42S}, \citet{2018arXiv180403143I}).  However, we do not often see multiple SMBHs at the centers of massive ellipticals (but see \citet{2016MNRAS.463.2145C}).  So, what does happen?  One paper (\citet{2018MNRAS.473.3410R}) has indicated that dynamical interactions among multiple orbiting black holes, which will eject a non-negligible fraction of the mass, may solve this problem.  The present paper also addresses this purported dynamical solution, focusing attention on the large fraction of lower mass black holes that remain to be detected as they orbit in massive galaxies.  What we argue in this paper is that, as a result of the N-body interactions among the infalling black holes, some are ejected, some remain in extended orbits and some (the few most massive ones) do in fact merge, but these do not exceed the PTA limits, and some of the orbiting ones might be detectable via their ability to accrete gas and emit radiation.

@@

In cosmological simulations it is common to overproduce stellar mass, with a general over-efficiency in the range of 2-4 times those measured in observations (\citet{1996ApJS..105...19K}, \citet{2010MNRAS.404.1111G}, \citet{2010ApJ...725.2312O}).  One typical reason cited for this overproduction is a lack or underestimation of supernova feedback effects in the form of thermal and/or kinetic energy.  However, even when thermal energy from supernova feedback is considered, the effects may be reduced due to the energy being quickly radiated away by rapidly cooling surrounding gas (\citet{1996ApJS..105...19K}).  A second common reason for overproduction of stars in cosmological simulations is a lack of modeling for AGN feedback, which may be responsible for solving the phenomenon called the "cooling flow paradox" (\citet{2001MNRAS.321L..20F}) that suppresses star formation.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{plots/stellar_to_halo_ratio.png}
\caption{Approximation of Figure 11 from \citet{2018AstL...44....8K}.  Purple dots shown are examples of our rescaled stellar masses}
\label{fig:stellar1}
\end{center}
\end{figure}

@@

\subsubsection{Kicked, Ejected and Orbiting Black Holes}\label{sec:kicked_orbiting}
Of the 37 orbiting black holes at $z=0$, 18 of them were kicked to higher orbits at some point in their traversal of their host galaxy.  If one counts the 13 ejected black holes, then of the 50 SMBHs that had not eventually merged with their host, 32 of them were kicked out of the galactic core as a result of three-body interactions.  If they did not merge, it was not for a lack of effort.

@@

\subsection{Central Black Hole Influence}
Within the influence radius of the central black hole, which is defined as the radius at which the total mass of the galaxy within that radius (stars + dark matter) equals the mass of the central black hole, the gas density is much higher than otherwise, and equation \eqref{beta_model} cannot be used in the region where the BH Kepler potential dominates.  Starting with the equation for hydrostatic equilibrium, assuming isothermality,
\begin{equation}
    C_{s}^2\frac{dln\rho}{dr}= -\frac{GM(r)}{r^2},
\end{equation}
we can integrate with respect to r, leading to 
\begin{equation}
    ln\frac{\rho}{\rho_1} = \frac{GM}{C_{s}^2}\left(\frac{1}{r} - \frac{1}{r_1}\right),
\end{equation}
where $\rho$ is the gas density at the location of the wandering black hole, $\rho_1$ is the density at the radius where the central black hole mass equals the galaxy mass, $r$ is the distance of the wandering black hole from the center, and $r_1$ is the radius where the central black hole mass equals the galaxy mass.

@@

\section{Conclusions}\label{sec:conclusions}
We performed N-body simulations on 11 galaxies ranging in mass between $10^{12} \msun < M < 10^{13} \msun$, from $4 \ge z \ge 0$.  We found that $42\%$ of the orbiting black holes merged with their host galaxy, while another $43\%$ remained in orbit at $z=0$.  With an x-ray luminosity range of $10^{25}-10^{43}$ $erg\ s^{-1}$
